---
title: "Reproducible Data - Course Project 1"
output: html_document
---
**1. Load and preprocess the data**
```{r echo=TRUE}
steps_df <- read.csv("activity.csv", na.strings="NA", stringsAsFactors=FALSE)
```

Note: this file did not require any other preprocessing before proceeding.



**2. What is the mean total number of steps taken per day?**


```{r echo=TRUE}
#Calculate the total steps per day and create a histogram of the results:

#get total number of steps per day
day_steps <- aggregate(steps ~ date, data = steps_df, FUN = sum)

#histogram it
hist(day_steps$steps, main = "Total Number of Steps per Day", 
     xlab = "Number of Steps", col = "red")
```

The mean number of total steps per day is `r sprintf("%.0f",mean(day_steps$steps, na.rm=TRUE),0)` steps, while the median is `r median(day_steps$steps, na.rm=TRUE)` per day.



**3. What is the average daily activity pattern?**

```{r echo=TRUE}
#Make a time series plot of the interval and avg steps across all days:

#get avg steps taken per interval
int_steps <- aggregate(steps ~ interval, data = steps_df, FUN = mean)

#plot the data
plot(int_steps, type="l", xlab = "Interval", ylab = "Average Steps", 
     main = "Average Steps per 5-Minute Interval" )
```

On average, the `r int_steps$interval[which.max(int_steps$steps)]`th 5-minute interval contains the maximum number of steps.


**4. Imputing missing values**

There are `r sum(is.na(steps_df))` missing values in this dataset. I will fill the missing values with the mean number of steps across all days for the particular interval.

```{r echo=TRUE}
#Create a new data set with missing values filled in:

imp_steps <- steps_df

for(r in 1:nrow(imp_steps)){
    if (is.na(imp_steps$steps[r])) {
        repl <- int_steps$steps[int_steps$interval == imp_steps$interval[r]];
        imp_steps$steps[r] <- repl;
    }
}
```

```{r echo=TRUE}
#Make a histogram of total steps per day with missing values filled in:

#get # of steps per day
day_steps <- with(imp_steps, tapply(steps, date, sum, na.rm=TRUE))

#histogram it
hist(day_steps, main = "Histogram of Number of Steps Per Day\nImputed Missing Values", 
     xlab = "Number of Steps Per Day", col = "red", breaks=15)
```

With the missing values filled in, the mean number of total steps per day is `r sprintf("%.0f",mean(day_steps))` steps, while the median is `r sprintf("%.0f",median(day_steps))` per day. The median did not change from the missing value data set, but the mean did go up one, a 0.00009 difference.  Filling missing values with the average of the corresponding interval had no real effect.

**5. Are there differences in activity patterns between weekdays and weekends?**

```{r echo=TRUE}
#Create a new factor variable in the dataset indicating if the day is a weekend or weekday:

#convert date field to Date datatype
imp_steps$date2 <- as.Date(imp_steps$date, format = "%Y-%m-%d")

#get day of week for each date
imp_steps$day <- weekdays(imp_steps$date2)

#set new column indicating if the day is a weekend or weekday
for (i in 1:nrow(imp_steps)) {
    if(imp_steps$day[i] == "Saturday" | imp_steps$day[i] == "Sunday") {
        imp_steps$day_type[i] <- "Weekend"
    } else {
        imp_steps$day_type[i] <- "Weekday"
    }
}

#convert to a factor for later
imp_steps$day_type <- factor(imp_steps$day_type)

```

```{r echo=TRUE}
#Make a time series plot of the average steps for each interval, divided out by weekend/weekday:

day_type_steps <- aggregate(steps ~ interval + day_type, data = imp_steps, FUN = mean)

library("lattice")

xyplot(steps ~ interval | day_type, day_type_steps, type = "l", layout = c(1, 2), 
       xlab = "Interval", ylab = "Number of Steps", 
       main = "Average Number of Steps per 5 Minute Interval\nWeekends vs Weekdays ")
